Random Walks across Social Networks: The Markovian Movement of a Stochastic Soccerball

### 2020-07-24 rmd/draft.Rmd
done almost

jupyter:
  jupytext:
    formats: nb//ipynb,rmd//Rmd
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.4.2
  kernelspec:
    display_name: R
    language: R
    name: ir
---

@@

Association football is the world's most popular game and a multibillion dollar industry. Statistical analysis of the sport was pioneered more than fifty years ago, but has only recently begun to make a more widespread impact. In this paper I explain how Social Network Analysis can provide the tools to model a football team's ball-possession as a dynamic probabilistic system, and hence how we can quantify the effect of each player on the team's goal-scoring capability.

@@

In this paper, I develop a rigorous account of a football match as a stochastic social network, the graph of which is generated by the movement of the ball from player to player. I then show how the adjacency matrix of a player, considered as a node in a football match's possession sequence graph, can be used to quantify his contribution to the likelihood of the team scoring. Finally, I discuss the potential and pitfalls of this new metric, which I call *Relative Markovian Expected Goals*.

@@

The dataset is provided by WyScout, one of a growing number of companies in the football analytics industry, and is available from FigShare.com [@LPappalardoMassucco2020]. It includes more than three million *event-logs* for every match from the five major European leagues, in JSON format (see Figure 1), with  the details of players, teams, and informative tags (such as the crucial 'Goal') numerically encoded. Each event is associated with an event type (such as 'Shot', 'Pass' or 'Free Kick'), identified with a player, and includes a nested array of the locational *positions* of the event itself and its target.

@@

We can then visualize the location of the event-log data points in the context of a diagrammatic football pitch, to allow for an immediate and accessible view of any particular possession sequence (eg. Table 2, Figure 3, Figure 4).

Figure 3: Pitchmap showing Possession Sequence
<img src="../figures/PossessionSequence.png">

We can compare this to a tabulated account of all events from just before the beginning of the possession sequence until its end, we see that our algorithm has successfully handled the case where an opposition player's unsuccessful challenge interrupts the sequence of events associated with the team in possession.

Table 2: Events of Possession Sequence

And with possession sequences that end in a goal, it is generally possible to find video highlights of the events under consideration, and confirm that the data we have does correspond to what actually happened.

Figure 4: YouTube Clip showing successful conclusion of possession sequence
<img src="../figures/youtube_clip.gif">

@@

Social network analysis applies the concepts of mathematical *graph theory*.

@@

A weighted simple graph can be uniquely described by an *adjacency matrix*:
$$
M_{ij} = \begin{cases}
    \begin{array}{ll}
         w(e_{i,j}), & e_{i,j} \in E \\
       0, & \mathrm{otherwise}
    \end{array}
\end{cases}
$$

@@

Figure 5
<img src="../figures/PossessionGraph.png">

There are several different 'weights' we could assign an edge: the Euclidean distance moved by the ball; the polar angle; the time interval. However, for this particular analysis we ignore the spatial dynamics and the subtler aspects of continuous time, and focus on the network connections between players.

@@

Figure 6
<img src="../figures/AdjacencyMatrixHome.png">

It is increasingly common [@TKnutson2018] to visualize a possession network by positioning each node according to the mean location of the player's touches, and varying the appearance (width, colour or transparency) of the edges to show the frequency of that pass. This swiftly communicates a team's formation, and gives a qualitative sense of a team's strategy.
Figure 7
<img src="../figures/PassingNetworkHome.png">

@@

A *random variable* is a function $X: \Omega \mapsto S$. If $S \subseteq \mathbb{R}$ then we also have the *expectation* $\mathbb{E}[X]$ and *variance* $\mathrm{Var}[X]$, but it is acceptable to have an $S \not\subset \mathbb{R}$. Again, to illustrate what we mean -- the number of goals in the match could be considered as a real-valued random variable, and we could talk about the expectation and variation of the number of goals; but this is not the case for the question of which player has the ball at a particular time, though this also is a random variable.

@@

Figure 8
<img src="../figures/TransitionMatrixHome.png">

@@

@@

One apparent flaw of this analysis is that it pays no attention to the *spatial* dynamics of the game. But it would be fairly straightforward to develop the analysis articulated here to also take into account a player's location on the pitch when he plays the ball. Rather than simply representing the players, the nodes' state space could instead be the Cartesian product of players with some set of discrete areas of the pitch, say $\{\Delta, \Phi, \Xi \}$ could represent the Defensive Half, Attacking Half, and Attacking Penalty Area respectively. One could simply add the spatial indicator to the player name at the preprocessing stage, and the rest of the analysis would then proceed as described here.

But the major weakness is one inherent in the data itself -- although the event-log dataset offers a wealth of information when compared to the manual *notational analysis* first developed by Charles Reep in the 1950s [@RPollard2002], a convincing analysis of football will only really be possible with data that includes the positions of off-the-ball players, as well as on-the-ball events. Such data can be captured [eg. @SPettersenEtAl2014], but is not yet widely (at least openly) available.

## Acknowledgments

@@

## Appendix 1: Analysis in R

>>>>

jupyter:
  jupytext:
    formats: nb//ipynb,rmd//Rmd
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.4.2
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
---

<small>

@@

Association football is the world's most popular sport [@IPalacios-Huerta2004] and a multibillion dollar industry [@Deloitte2019]. 
Statistical analysis of the sport was pioneered more than fifty years ago [@CReepBenjamin1968], but has only recently begun to make a more widespread impact [@PCintiaEtAl2015]. Such analysis is of potential interest to coaches and scouts working professionally within the game, journalists and media commentators explaining the game, and gamblers and bookmakers putting money on the game. In this paper I explain how Social Network Analysis [@SWassermanEtAl1994] can provide the tools to model a football team's ball-possession as a dynamic probabilistic system, and hence how we can quantify the contribution of each player to the team's goal-scoring capability.

</small>

@@

In this paper, I develop a rigorous account of a football match as a stochastic social network, the graph of which is generated by the movement of the ball from player to player. I then show how the adjacency matrix of a player, considered as a node in a football match's possession sequence graph, can be used to quantify his contribution to the likelihood of the team scoring, which I call *Markov Expected Goals*. Finally, I discuss the potential and pitfalls of this new metric.

@@

The dataset is provided by WyScout, one of a growing number of companies in the football analytics industry, and is available from FigShare.com [@LPappalardoMassucco2019]. It includes more than three million *event-logs* for every 2017-18 match from the five major European leagues, in JSON format (see Figure 1), with  the details of players, teams, and informative tags (such as the crucial 'Goal') numerically encoded. Each event is associated with an event type (such as 'Shot', 'Pass' or 'Free Kick'), identified with a player, and includes a nested array of the locational *positions* of the event itself and its target.

<details open>
<summary>Figure 1</summary>
<figure>
----

```{python echo=False, wrap=False}

with open('../figures/ExampleEvent.json','r') as f:
	json_event = f.read()
print(json_event)

```

<figcaption>
Figure 1: Example of Event-Log in JSON Format
</figcaption>
----

</figure>
</details>

@@

We can then visualize the location of the event-log data points in the context of a diagrammatic football pitch, to allow for an immediate and accessible view of the events of any match (see Figures 2 & 3), or particular possession sequence (see Figure 4).

<details open>
<summary>Table 1</summary>
<figure>
----

<figcaption>
Table 1: Frequency of Event Types
</figcaption>

```{python echo=False, wrap=False, results='table'}

import pandas as pd

table1 = pd.read_csv('../tables/EventTypeTally.csv',
	index_col=0,dtype=str)
print(table1.to_markdown(floatfmt="#.3g"))

```

----

</figure>
</details>

<details open>
<summary>Table 2</summary>
<figure>
----

<figcaption>
Table 2: Frequency of Pass Types
</figcaption>

```{python echo=False, wrap=False, results='table'}

import pandas as pd

table1 = pd.read_csv('../tables/PassTally.csv',
	index_col=0,dtype=str)
print(table1.T.to_markdown(floatfmt="#.3g"))

```

----

</figure>
</details>

<details open>
<summary>Figure 2</summary>
<figure>
----

<table><tr><td>
<img src="../figures/EventMapHome.png">
</td><td>
<img src="../figures/EventMapAway.png">
</td></tr></table>

<figcaption>
Figure 2: Events for Home and Away Team
</figcaption>
----

</figure>
</details>

<details open>
<summary>Figure 3</summary>
<figure>
----

<table><tr><td>
<img src="../figures/PassMapHome.png">
</td><td>
<img src="../figures/PassMapAway.png">
</td></tr></table>

<figcaption>
Figure 3: Passes and Shots for Home and Away Team
</figcaption>

----

</figure>
</details>

<details open>
<summary>Figure 4</summary>
<figure>
----

<img src="../figures/PossessionSequence.png">
<figcaption>Figure 4: Pitchmap showing Possession Sequence</figcaption>

----

</figure>
</details>

We can compare this to a tabulated account of all events from just before the beginning of the possession sequence until its end, and we see that our algorithm has successfully handled the case where an opposition player's unsuccessful challenge interrupts the sequence of events associated with the team in possession.

<details open>
<summary>Table 1</summary>
<figure>
----

<figcaption>
Table 1: Events of Possession Sequence
</figcaption>

```{python echo=False, wrap=False, results='table'}

import pandas as pd

table1 = pd.read_csv('../tables/1.csv',index_col=0)
print(table1.to_markdown(showindex=False))

```

----

</figure>
</details>

With possession sequences that end in a goal, it is generally possible to find video highlights of the events under consideration, and confirm that the data we have does correspond to what actually happened.

<details open>
<summary>Figure 5</summary>
<figure>
----

<img src="../figures/YouTubeClip.gif" width=800>
<figcaption>Figure 5: YouTube Clip showing successful conclusion of possession sequence
</figcaption>

----

</figure>
</details>

@@

Social Network Analysis applies the concepts of mathematical *graph theory*.

@@

A weighted simple graph can be uniquely described by an *adjacency matrix*:
$$
M_{ij} = \begin{cases}
    \begin{array}{ll}
         w(e_{i,j}), & e_{i,j} \in E \\
       0, & \mathrm{otherwise.}
    \end{array}
\end{cases}
$$

@@

<details open>
<summary>Figure 6</summary>
<figure>
----

<img src="../figures/PossessionGraph.png">
<figcaption>Figure 6: Graph of Possession Sequence leading to Goal
</figcaption>

----

</figure>
</details>

There are several different 'weights' we could assign an edge: the Euclidean distance moved by the ball; the polar angle; the time interval. However, for this particular analysis we focus simply on the network connections between players.

@@

<details open>
<summary>Figure 7</summary>
<figure>
----

<img src="../figures/AdjacencyMatrixAway.png">

<figcaption>
Figure 7: Heatmap of Adjacency Matrix
</figcaption>

----

</figure>
</details>

It is increasingly common [@TKnutson2018] to visualize a possession network by positioning each node according to the mean location of the player's touches, and varying the appearance (width, colour or transparency) of the edges to show the frequency of that pass. This swiftly communicates a team's formation, and gives a qualitative sense of a team's strategy.

<details open>
<summary>Figure 8</summary>
<figure>
----

<table><tr><td>
<img src="../figures/PassingNetworkHome.png">
</td><td>
<img src="../figures/PassingNetworkAway.png">
</td></tr></table>

<figcaption>
Figure 8: Possession Networks for Home and Away Team
</figcaption>

----

</figure>
</details>

@@

A *random variable* is a function $X: \Omega \mapsto S$. If $S \subseteq \mathbb{R}$ then we also have the *expectation* $\mathbb{E}[X]$ and *variance* $\mathrm{Var}[X]$. Again, to illustrate what we mean -- the number of goals in the match is an integer-valued random variable, and since $\mathbb{Z} \subset \mathbb{R}$ we could talk about the expectation and variation of the number of goals; but this is not the case for the question of which player has the ball at a particular time, though this also is a random variable.

@@

<details open>
<summary>Figure 9</summary>
<figure>
----

<img src="../figures/TransitionMatrixAway.png">

<figcaption>
Figure 9: Heatmap of Transition Matrix
</figcaption>

----

</figure>
</details>

@@

<details open>
<summary>Table 2</summary>
<figure>
----

<figcaption>
Table 2: Top Twenty European Players by MXG Score
</figcaption>

```{python echo=False, wrap=False, results='table'}

import pandas as pd

MXGtable = pd.read_csv('../tables/2.csv',index_col=0)
print(MXGtable.to_markdown(floatfmt="#.3g"))

```

----

</figure>
</details>

@@

One apparent flaw of this analysis is that it pays no attention to the *spatial* dynamics of the game. But it would be fairly straightforward to develop the analysis articulated here to also take into account a player's location on the pitch when he plays the ball. Rather than merely representing the players, the nodes' state space could instead be the Cartesian product of players with some set of discrete areas of the pitch, say $\{\Delta, \Phi, \Xi \}$ representing the Defensive Half, Attacking Half, and Attacking Penalty Area respectively. One could simply add the spatial indicator to the player name at the preprocessing stage, and the rest of the analysis would then proceed as described here.

But the major weakness is one inherent in the data itself -- although the event-log dataset offers a wealth of information when compared to the manual *notational analysis* first developed by Charles Reep in the 1950s [@RPollard2002], a convincing analysis of football will only really be possible with data that includes the positions of off-the-ball players, as well as on-the-ball events [@RTavares2017]. Such data can be captured [eg. @SPettersenEtAl2014], but is not yet widely (at least openly) available.

## Appendix 1: Social Network Analysis in R

<details open>
<summary>Code</summary>

```{python echo=False, wrap=False, results='Rmd'}

with open('../rmd/SocialNetworkAnalysis_R.Rmd','r') as f:
	R_script = f.read()
print(R_script)

```

</details>

<details open>
<summary>Code</summary>

```{python echo=False, wrap=False, results='Rmd'}

with open('../rmd/Preprocessing_Python.Rmd','r') as f:
	python_script = f.read()
print(python_script)

```

</details>

@@
